prolfqua - A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-06-03 inst/MyArticle/paper.Rmd
fix references by PN.

---
title: prolfqua - A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [Swiss Federal Institute of Technology in Zurich](https://ethz.ch) \| [University of Zurich](https://uzh.ch); Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB)](https://www.sib.swiss/) $^\dagger$Correspondence \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
      keep_tex: true
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
    Protein mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. Nevertheless, there is a need for a flexible and easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package can model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. It integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The application programmer interface strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis easy and exciting. 
     Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. Finally, we show that the implemented methods allow sensitive and specific differential expression analysis. 
     The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT licence, and runs on all plattforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).
urlcolor: blue
---

@@

An important aspect of mass spectrometric data are missing peptide and protein quantifications. [@rubin1976inference] classified missing data problems into three categories: missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). For instance, in data-dependent acquisition (DDA) MS, only a limited number of MS1 signals are selected for fragmentation and identified. Peptide quantification algorithms transfer identification information between MS1 features in different samples by aligning the data using retention time and mass information, thus reducing the amount of missing data. Nevertheless, highly abundant proteins can suppress the detection of other proteins in a sample, an MAR mechanism. Furthermore, a weak correlation between the number of missing measurements in a condition and the abundance of the quantified protein can be observed, which is caused by the limit of detection (LOD), an MNAR mechanism [@mcgurk2020use].

Several packages exist to model mass spectrometry protein quantification experiments, e.g., `r BiocStyle::Biocpkg("limma")` [@Ritchie2015], `r BiocStyle::Biocpkg("MSstats")` [@Choi2014], `r BiocStyle::Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r BiocStyle::Biocpkg("msqrob2")` [@Goeminne2016] or `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in the R. Each of them has some unique features, for example, `r BiocStyle::Biocpkg("MSstats")` infers the model and generates the model formula from the sample annotation, allowing users with limited statistical knowledge to perform differential expression analysis (DEA). At the same time, `r BiocStyle::Biocpkg("limma")` allows to specify a design matrix using a linear model formula and implements the empirical Bayes variance shrinkage method. The package `r BiocStyle::Biocpkg("PECA")`, performs a roll-up of peptide level differences and peptide level $p$-value estimates, obtained from `r BiocStyle::Biocpkg("limma")` or `r BiocStyle::Biocpkg("PECA")`, to protein level estimates. Furthermore, `r BiocStyle::Biocpkg("msqrob2")` implements two models: robust linear models fitted to protein intensities and robust ridge regression fitted to peptide intensities and combines them with empirical Bayes variance shrinkage. The `r BiocStyle::Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation. `r BiocStyle::Biocpkg("MSstats")` handles missing data by feature filtering and using imputation. Other means of modelling missing observations are the Hurdle model discussed by [@goeminne2020msqrob], while the R package `r BiocStyle::Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020].

@@

Valikangas [@valikangas2016systematic] and colleagues discuss and benchmark various methods of peptide or protein intensity normalization, such as variance stabilizing normalization [@huber2002variance] or quantile normalization [@bolstad2003comparison]. In this work, we will be using a robust version of the z-score, where instead of the mean $\bar{x}$ we use the median, and instead of the standard deviation $\tilde{S}$ the median absolute deviation (mad):

@@

From the linear and the mixed effect models, we can obtain the residual standard deviation $\sigma$, and degrees of freedom $df$. [@Smyth2004linear] discuss how, to use the $\sigma$ and $df$ of all models to estimate a prior $\sigma$ and prior $df$, and posterior $\tilde \sigma$. These can be used to moderate the $t$-statistics by:

>>>>

---
title: prolfqua - A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [Swiss Federal Institute of Technology in Zurich](https://ethz.ch) \| [University of Zurich](https://uzh.ch); Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB)](https://www.sib.swiss/) $^\dagger$Correspondence \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. Nevertheless, there is a need for a flexible and easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package can model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. It integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The application programmer interface strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis easy and exciting. 
     Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. Finally, we show that the implemented methods allow sensitive and specific differential expression analysis. 
     The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT licence, and runs on all plattforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).
urlcolor: blue
---

@@

An important aspect of mass spectrometric data are missing peptide and protein quantifications. @rubin1976inference classified missing data problems into three categories: missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). For instance, in data-dependent acquisition (DDA) MS, only a limited number of MS1 signals are selected for fragmentation and identified. Peptide quantification algorithms transfer identification information between MS1 features in different samples by aligning the data using retention time and mass information, thus reducing the amount of missing data. Nevertheless, highly abundant proteins can suppress the detection of other proteins in a sample, an MAR mechanism. Furthermore, a weak correlation between the number of missing measurements in a condition and the abundance of the quantified protein can be observed, which is caused by the limit of detection (LOD), an MNAR mechanism [@mcgurk2020use].

Several packages exist to model mass spectrometry protein quantification experiments, e.g., `r BiocStyle::Biocpkg("limma")` [@Ritchie2015], `r BiocStyle::Biocpkg("MSstats")` [@Choi2014], `r BiocStyle::Biocpkg("PECA")` [@Suomi2017bEnhanced],  `r BiocStyle::Biocpkg("msqrob2")` [@Goeminne2016] or `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020], to mention some, all implemented in the R. Each of them has some unique features, for example, `r BiocStyle::Biocpkg("MSstats")` infers the model and generates the model formula from the sample annotation, allowing users with limited statistical knowledge to perform differential expression analysis (DEA). At the same time, `r BiocStyle::Biocpkg("limma")` allows to specify a design matrix using a linear model formula and implements the empirical Bayes variance shrinkage method. The package `r BiocStyle::Biocpkg("PECA")`, performs a roll-up of peptide level differences and peptide level $p$-value estimates, obtained from `r BiocStyle::Biocpkg("limma")` or `r BiocStyle::Biocpkg("PECA")`, to protein level estimates. Furthermore, `r BiocStyle::Biocpkg("msqrob2")` implements two models: robust linear models fitted to protein intensities and robust ridge regression fitted to peptide intensities and combines them with empirical Bayes variance shrinkage. The `r BiocStyle::Biocpkg("proDA")` package implements a linear probabilistic dropout model to account for missing data without imputation. `r BiocStyle::Biocpkg("MSstats")` handles missing data by feature filtering and using imputation. Other means of modelling missing observations are the Hurdle model discussed by @goeminne2020msqrob, while the R package `r BiocStyle::Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020].

@@

@valikangas2016systematic discuss and benchmark various methods of peptide or protein intensity normalization, such as variance stabilizing normalization [@huber2002variance] or quantile normalization [@bolstad2003comparison]. In this work, we will be using a robust version of the z-score, where instead of the mean $\bar{x}$ we use the median, and instead of the standard deviation $\tilde{S}$ the median absolute deviation (mad):

@@

From the linear and the mixed effect models, we can obtain the residual standard deviation $\sigma$, and degrees of freedom $df$. @Smyth2004linear discuss how, to use the $\sigma$ and $df$ of all models to estimate a prior $\sigma$ and prior $df$, and posterior $\tilde \sigma$. These can be used to moderate the $t$-statistics by:
